\textbf{Are regime changes always bad for markets? Evidence from daily financial data}

### 2021-11-16 doc/regime_changes_note.tex
Additional edits to note draft

Economies are now global, financialized, and integrated, with domestic economies intrinsically linked to international investment. Political instability is negatively correlated with investment, financial development, and GDP growth \citep{aisen2013does, alesina1996income, alesina1996political, fosu1992political, jong2009measurement, roe2011political, baker2016measuring}, as well as associated with increases in stock market variance \citep{leblang2005government, jensen2005market, liu2015economic}.

@@

However, how different types of political instability affect a country's access to capital remains unknown. In contrast to the aforementioned cross country studies, we therefore separately examine the effect of various types of instability (coups, resignations, assassinations, and protests) on domestic access to financial capital. Specifically, we examine whether changes in financial capital flows differ for coups, resignations, assassinations, and protests, as well as for authoritarian vs. democratic and pro vs. anti business shifts.

To test our concept, we examine shifts in capital market flows surrounding politically unstable domestic events. Specifically, we conduct event studies of daily financial data, which estimate a local average treatment effect of an unexpected event on stock prices \textit{at the time of the event}. This interrupted time-series approach mitigates the endogeneity problems in previous cross country regressions---confounding events would need to occur on the same day as instability, and do so for a large portion of all of our independently tested events in order to influence our estimates.

@@

We analyze the full sample of politically unstable events for which national-level daily financial data is available.\footnote{13 coups, 8 assassinations, 15 forced resignations, and 11 public protests.}  Like previous research, we find that all types of political instability in our sample cause large increases in financial volatility. However, we find that capital flows are large and positive (+4\%) following resignations and negative and smaller in magnitude following assassinations and coups (-2\%). We also find that the failed pro-business 2002 Venezuelan coup caused large positive capital inflows (+10\%), followed by immediate capital flight (-8\%) following the reinstatement of a left-wing populist.

Our primary contributions are empirical and methodological. Empirically, we provide the first estimates of capital market reactions disaggregated by type of political instability, and show that instability does not systematically depress investment. The capital flows we document are not insubstantial---in many cases they represent larger shocks than the 2008 stock market crash on their respective domestic economies. Methodologically, we (1) employ a method less susceptible to endogeneity concerns than previous studies, and (2) integrate synthetic control and event study methods to allow for control portfolios when a control candidate is not present. Theoretically, we find evidence that authoritarian regime changes are more likely to lead to negative returns, but that leaders who are clearly pro-business can be rewarded by financial markets, even if they use extra-judicial methods to take power.

@@

Conventional theory and cross-country empirical evidence suggests that political instability depresses financial returns and economic growth \citep{irshad2017relationship, le2006political, lensink2000capital, boutchkova2012precarious, lehkonen2015democracy}. But if market returns reflect investor expectations about future economic growth and returns, such hypotheses may be too simplistic. Investors may view even a coup as a ``positive'' event if the current regime is anti-business or anti-global.

@@

Not all regime changes are equivalent. Some regime changes---e.g., the resignation of an ineffective leader---may foreshadow better policy. Assassinations can occur seemingly at random, and a successor may be unclear. Coups can be democratic or autocratic. These differences are often overlooked in past studies, which have proxied for instability using the number of coups \citep{londregan1990poverty, alesina1996political}, assassinations or revolutions \citep{barro1991economic}, or combined events into single indices \citep{alesina1996income, venieris1986income, gupta1990economics, jong2009measurement}. By contrast, we estimate effects separately for coups, assassinations, resignations, and protests.

@@

Debate also exists on the subject of ``good coups''---i.e., coups that lead to democratization or economic liberalization---and their effect on economic growth.  Most research suggests ``good coups'' are not the norm \citep{derpanopoulos2015coups, powell2011global, thyne2016coup, varol2011democratic}. Others argue the negative effects of uncertainty dominate any positive effects \citep{alesina1996political}, or that the expected impact of coups is neutral as some enhance growth while others depress it \citep{londregan1990poverty}. But while ``good coups'' may be the minority, they have become more frequent \citep{marinov2014coups} and could have positive economic effects \citep{meyersson2016political, girardi2018institution}.

@@

Previous research therefore suggests that coups should on average cause negative financial flows, but sometimes lead to positive flows when the coup's instigators are more pro-market than the regime they replace.  Assassinations should have a neutral or negative impact as they increase uncertainty, but institutional responses to assassinations vary by country. By contrast, resignations may be viewed positively on average, as they typically signal the departure of an ineffective leader.

@@

To estimate the effect of irregular regime changes on financial flows, we follow standard event study methodology \citep{mackinlay1997event, campbell1997econometrics}. Normal performance is measured with a constant mean return model,
\begin{align} \label{eqn:market-model}
R_{it}=\mu_{i}+\epsilon_{it},
\end{align}
where $R_{it}$ is the logged return of national stock index $i$ on trading day $t$ and $\epsilon_{it}$ is the error term. We calculate abnormal returns (ARs) in an ``event window'' around the date of each event, $AR_{i\tau}=R_{i\tau}-\widehat{\mu}_i$, where $\tau$ is a date in the event window, and $\widehat{\mu}_i$ is estimated in an ``estimation window'' preceding the event window with \autoref{eqn:market-model}. We use a 41 trading day event window (20 pre-event days, event day, and 20 post-event days) and 250 trading day estimation window. The ARs are then used to calculate cumulative abnormal returns (CARs) between event day $\tau_1$ and event day $\tau_2$: $CAR(\tau_1,\tau_2)=\sum_{\tau=\tau_1}^{\tau_2}AR_{i\tau}$. Standard errors and p-values are calculated using asymptotic t-statistics as in \citet{mackinlay1997event}.\footnote{It is appropriate to use the standard normal distribution to calculate test statistics because the length of the estimation window is sufficiently long (250 trading days).} The event date is the first trading day a market could react to news of the event.

@@

We do not use a market model (where a market index is included as a control) because our unit of analysis \textit{is} the country-wide market index (not firms).\footnote{Plausible market indices such as the MSCI World Index and the S\&P/IFC Emerging Markets Investable Composite Index only begin in 1976 and 1995, respectively.} To address concerns regarding use of a constant mean return model, we combine synthetic control methods \citep{abadie2003economic, abadie2010synthetic} with event study estimation. We create a ``synthetic'' control portfolio for each event, where each country is given a weight representing its influence in the control portfolio. The weight is chosen so that the daily returns and the variance of the daily returns of the control portfolio and the event country are most similar in the estimation window. The set of possible countries in the control portfolio consists of all countries listed in \autoref{tab:stock-list}. A more formal exposition of the methodology can be found in \nameref{subsec: synth}.

@@

We first confirm that our sample of events shows the same increases in volatility suggested by previous literature. \autoref{fig:volatility} shows the mean volatility ($\overline{\sigma_t}$) estimates across all irregular regime changes for 250 trading days prior to and 250 days after each event. Volatility stays between a narrow range at nearly all dates except those surrounding the regime change. There is an enormous volatility jump on the day of the regime change, with levels then decreasing to normal within a month of the event.

@@

The average coup has a $-2.1\%$ event day AR. Event day ARs for the 1970 coup in Argentina, the 1991 coup in Thailand, the 1992 coup in Peru, and the 1999 coup in Pakistan are all negative and statistically different than zero. Moreover, all of these cases except Thailand have negative pre-event CARs that are statistically indistinguishable from zero, suggesting they were not foreseen. In all of these cases, the coup either overthrew a democratically elected government or changed governance from one military ruler to another. The initial negative reaction followed by additional post-event negativity is consistent with the expected market reaction from a successful authoritarian coup followed by post-event consolidation of power.

@@

The only events with positive ARs are the 1971 coup in Argentina and the 2002 coup in Nepal. These results provide evidence that coups do not necessarily lead to negative abnormal returns. While the 1971 Argentinian coup did result in another military leader, it did so while calling for free and democratic elections and replaced a government that had adopted extreme protectionist economic policies. In fact, by 1973 Argentina had transitioned to a democracy.\footnote{Based on Center for System Peace Polity IV polity score of 6. Values of 6-10 are defined as democracies.} The 2002 coup in Nepal resulted in a monarchical restoration, but occurred after the country's prime minister postponed general elections, itself a democratically subversive action.

@@

Like the majority of coups, we find that assassinations decrease stock prices (see \autoref{fig:mean-car-by-regime-change} and \autoref{tab:AR-coups}). The mean event day abnormal return is negative and statistically different than zero.

@@

These results are consistent with our hypothesis that resignations may lead to positive returns as they typically occur due to poor performance and/or loss of authority. Among our sample of events, leaders were ousted following loss in conflict/war, anti-authoritarian protest, corruption scandals, Supreme Court ruling against unconstitutional actions, contested elections, and abuse of power.

@@

The resignations analyzed encompass leaders who left office because of poor performance, public discontent and popular protests. Protests preceding these resignations may therefore also affect financial markets\footnote{Indeed, corporate investors in the 2013 MIGA \textit{World Investment and Political Risk} ranked civil disturbances as the fourth most concerning type of political risk.}, so we analyze all protests that preceded the resignations in \nameref{subsec: Public Protests}. Including directionality, public protests appear to have no effect on stock returns as some increase stock prices while others decrease them (see \autoref{tab:protest-stocks}), but in absolute terms,  returns are approximately 1.5\% higher during public protests.

@@

We therefore aggregate the events in our sample by whether they resulted in an authoritarian or democratic shift.\footnote{As defined by the Polity project.} We find suggestive evidence that regime changes associated with authoritarian shifts are on average perceived negatively by investors (see \autoref{tab:AR-auth}), while democratic shifts are perceived positively (see \autoref{tab:AR-dem}). Of ten authoritarian shifts, only two result in positive CARs, and neither are significantly different from zero. Three of five cases involving democratic shifts result in positive CARs, and the positive CARs persist in the follows weeks. Of the two negative CARs, one is not significantly different from zero, and the other is associated with a forced market closure that lasted 17 days.

@@

\autoref{fig:AR-Ven} shows CARs and 95\% confidence intervals for the 20 days surrounding the event. The abnormal return on the first trading day investors could react to the coup was +10\%. When Chavez was reinstated, the abnormal return was -8\%. The 0\% 10-day CAR preceding the coup implies that investors were completely unaware of the coup plot, increasing our confidence that the abnormal returns capture the true effect of the regime changes. This failed coup demonstrates a large positive market reaction to the attempted overthrow of a socialist leader, and an equally large negative reaction to his reinstatement.  The large magnitudes and precision of these effects suggest that investors value transition to a pro-business government regime, regardless of how the regime change is achieved.

@@

Conventional theory suggests political instability causes capital flight. We show that this is not always the case. Unexpected changes in ruler virtually always increase volatility, but returns are not always negative. Capital markets can be given a boost when a new regime is expected to offer a more democratic or pro-business environment.

Despite their frequency, there is little evidence on the economic consequences of various types of political instability.  We show that market returns following resignations are large and positive on average, but negative and smaller in magnitude following assassinations and coups. We also find evidence that capital markets tend to prefer democratic regime changes to authoritarian shifts, but that even democratically subversive coups can boost markets if the instigators are clearly pro-business.

There are a number of avenues for future research. First, more research is needed to identify the pro and anti market characteristics of regime changes. Second, more work is needed to determine the extent to which financial market flows translate to broader economic development outcomes. It remains unclear whether the direction of the effects of different types of regime changes on outcomes such as economic growth, investment, debt, inflation, infant mortality, and years of schooling are consistent with their stock market effects, or if investor perceptions are at odds with certain development goals.

>>>>

Economies are now global, financialized, and integrated, with domestic economies intrinsically linked to international investment. However, political instability appears to jeopardize access to capital---instability is negatively correlated with investment, financial development, and GDP growth \citep{aisen2013does, alesina1996income, alesina1996political, fosu1992political, jong2009measurement, roe2011political, baker2016measuring}, as well as associated with increases in stock market variance \citep{leblang2005government, jensen2005market, liu2015economic}.

@@

However, how different \textit{types} of political instability affect domestic capital access remains unclear. In contrast to the aforementioned cross country studies, we therefore separately examine the effect of different types of instability on domestic firms' access to financial capital. Specifically, we examine whether changes in financial flows differ for coups, resignations, assassinations, and protests, as well as for authoritarian vs. democratic and pro vs. anti business shifts.

To test our concept, we examine changes in market returns surrounding politically unstable domestic events. Specifically, we conduct event studies of daily financial data, which estimate a local average treatment effect of an unexpected event on stock prices \textit{at the time of the event}. This interrupted time-series approach mitigates the endogeneity problems in previous cross country regressions---confounding events would need to occur on the same day as instability, and do so for a large portion of all of our independently tested events in order to influence our estimates.

@@

We analyze the full sample of politically unstable events for which national-level daily financial data is available.\footnote{13 coups, 8 assassinations, 15 forced resignations, and 11 public protests.}  Like previous research, we find that all types of political instability in our sample cause large increases in financial volatility. However, we find that market returns are large and positive (+4\%) following resignations and negative and smaller in magnitude following assassinations and coups (-2\%). We also find that the failed pro-business 2002 Venezuelan coup caused large positive market returns (+10\%), followed by immediate capital flight (-8\%) following the reinstatement of a left-wing populist.

Our primary contributions are empirical and methodological. Empirically, we provide the first estimates of the effects of different types of political instability on domestic capital access, and show that instability does not systematically depress investment. The capital flows we document are not insubstantial---they are in many cases larger shocks than the 2008 stock market crash on their respective domestic economies. Methodologically, we (1) employ a method less susceptible to endogeneity concerns than previous studies, and (2) integrate synthetic control and event study methods to allow for control portfolios when a control candidate is not present. Theoretically, we find evidence that authoritarian regime changes are more likely to lead to negative returns, but that leaders who are clearly pro-business can be rewarded by financial markets, even if they use extra-judicial methods to take power.

@@

Conventional theory and cross-country empirical evidence suggests that political instability depresses financial returns and economic growth \citep{irshad2017relationship, le2006political, lensink2000capital, boutchkova2012precarious, lehkonen2015democracy}, and increases in volatility \citep{irshad2017relationship, bialkowski2008stock, leblang2005government, jensen2005market, liu2015economic}. But if capital flows reflect investor expectations about future economic growth and returns, such hypotheses may be too simplistic. For example, capital access may increase in response to a coup if the current regime is anti-business or anti-global.

@@

Not all irregular regime changes are equivalent. Some---e.g., the resignation of an ineffective leader---may foreshadow better policy. Assassinations can occur seemingly at random, and a successor may be unclear. Coups can be democratic or autocratic. These differences are often overlooked in past studies, which have proxied for instability using the number of coups \citep{londregan1990poverty, alesina1996political}, assassinations or revolutions \citep{barro1991economic}, or combined events into single indices \citep{alesina1996income, venieris1986income, gupta1990economics, jong2009measurement}.

@@

Debate also exists on ``good coups''---i.e., coups that lead to democratization or economic liberalization---and their effect on economic growth.  Most research suggests these are not the norm \citep{derpanopoulos2015coups, powell2011global, thyne2016coup, varol2011democratic}. Others argue the negative effects of uncertainty dominate any positive effects \citep{alesina1996political}, or that the expected impact of coups is neutral as some enhance growth while others depress it \citep{londregan1990poverty}. But while ``good coups'' may be the minority, they have become more frequent \citep{marinov2014coups} and could have positive economic effects \citep{meyersson2016political, girardi2018institution}.

@@

Previous research therefore suggests coups should on average cause capital flight, but lead to positive flows if the coup's instigators are more pro-market than the regime they replace.  Assassinations should have a neutral or negative impact as they increase uncertainty, but institutional responses to assassinations vary by country. By contrast, resignations may be viewed positively on average, as they often signal the departure of an ineffective leader.

@@

To estimate the effect of irregular regime changes on financial flows, we follow standard event study methodology \citep{mackinlay1997event, campbell1997econometrics}. Normal performance is measured with a constant mean return model,
\begin{align} \label{eqn:market-model}
R_{it}=\mu_{i}+\epsilon_{it},
\end{align}
where $R_{it}$ is the logged return of national stock index $i$ on trading day $t$ and $\epsilon_{it}$ is the error term. We calculate abnormal returns (ARs) in an ``event window'' around the date of each event, $AR_{i\tau}=R_{i\tau}-\widehat{\mu}_i$, where $\tau$ is a date in the event window, and $\widehat{\mu}_i$ is estimated in an ``estimation window'' preceding the event window with \autoref{eqn:market-model}. We use a 41 trading day event window (20 pre-event days, event day, and 20 post-event days) and 250 trading day estimation window. The ARs are then used to calculate cumulative abnormal returns (CARs) between event day $\tau_1$ and event day $\tau_2$: $CAR(\tau_1,\tau_2)=\sum_{\tau=\tau_1}^{\tau_2}AR_{i\tau}$. Standard errors and p-values are calculated using asymptotic t-statistics as in \citet{mackinlay1997event}.\footnote{This is appropriate because the length of the estimation window is sufficiently long (250 trading days).} The event date is the first trading day a market could react to news of the event.

@@

We do not use a market model (where a market index is included as a control) as our unit of analysis \textit{is} the country-wide market index (i.e., not a firm). To address concerns regarding use of a constant mean return model, we combine synthetic control methods \citep{abadie2010synthetic} with event study estimation. We create a ``synthetic'' control portfolio for each event, where each country is given a weight representing its influence in the portfolio. The weight is chosen so that the daily returns and the variance of the daily returns of the control portfolio and the event country are most similar in the estimation window. The possible countries in the control portfolio are all countries listed in \autoref{tab:stock-list}. A more formal exposition of the methodology can be found in \nameref{subsec: synth}.

This interrupted time series approach implies that the ``control group'' being compared against is not, for example, regular regime changes or failed regime changes, but the expected returns in the same country on the same day in a but-for world in which no regime change occurred. We report abnormal returns separately for coups, assassinations and resignations. However, each event CAR is by itself also a valid causal estimate of the effect of a specific regime change.

@@

We first confirm that our sample of events exhibits the increases in volatility suggested by previous literature. \autoref{fig:volatility} shows the mean volatility ($\overline{\sigma_t}$) estimates across all irregular regime changes for 250 trading days prior to and 250 days after each event. Volatility stays between a narrow range at nearly all dates except those surrounding the regime change. There is an enormous volatility jump on the day of the regime change, with levels then decreasing to normal within a month of the event.

@@

The average coup has a -2.1\% event day AR. ARs for the 1970 coup in Argentina,  1991 coup in Thailand, 1992 coup in Peru, and 1999 coup in Pakistan are all negative and significantly different than zero. Moreover, all of these cases except Thailand have negative pre-event CARs that are statistically indistinguishable from zero, suggesting they were not foreseen. In all of these cases, the coup either overthrew a democratically elected government or changed governance from one military ruler to another. This initial capital flight followed by additional post-event negativity is consistent with the expected market reaction from a successful authoritarian coup followed by post-event consolidation of power.

@@

The only events with positive ARs are the 1971 coup in Argentina and the 2002 coup in Nepal. These results provide evidence that coups do not necessarily lead to negative abnormal returns. While the 1971 Argentinian coup did result in another military leader, it did so while calling for free and democratic elections and replaced a government that had adopted extreme protectionist economic policies. In fact, by 1973 Argentina had transitioned to a democracy.\footnote{Based on Center for System Peace Polity IV polity score of 6. Values of 6-10 are defined as democracies.} The 2002 coup in Nepal resulted in a monarchical restoration, but occurred after the prime minister postponed general elections, itself a democratically subversive action.

@@

Like the majority of coups, we find that assassinations lead to capital flight (see \autoref{fig:mean-car-by-regime-change} and \autoref{tab:AR-coups}). The mean event day abnormal return is negative and statistically different than zero.

@@

These results are consistent with our hypothesis that resignations may lead to increased capital access as they typically occur due to poor performance and/or loss of authority. Among our sample of events, leaders were ousted following loss in conflict/war, anti-authoritarian protest, corruption scandals, Supreme Court ruling against unconstitutional actions, contested elections, and abuse of power.

@@

The resignations analyzed encompass leaders who left office because of poor performance, public discontent and popular protests. Protests preceding these resignations may therefore also affect capital access,\footnote{Indeed, corporate investors in the 2013 MIGA \textit{World Investment and Political Risk} ranked civil disturbances as the fourth most concerning type of political risk.} so we analyze all protests that preceded the resignations in \nameref{subsec: Public Protests}. Including directionality, public protests appear to have no effect on returns as some increase stock prices while others decrease them (see \autoref{tab:protest-stocks}), but in absolute terms, returns are approximately 1.5\% higher during public protests.

@@

We therefore aggregate the events in our sample by whether they resulted in an authoritarian or democratic shift.\footnote{As defined by the Polity project.} We find suggestive evidence that regime changes associated with authoritarian shifts are on average perceived negatively by investors (see \autoref{tab:AR-auth}), while democratic shifts are perceived positively (see \autoref{tab:AR-dem}). Of ten authoritarian shifts, only two result in positive CARs, and neither are significantly different from zero. Three of five cases involving democratic shifts result in positive CARs, and the positive CARs persist in the follows weeks. Of the two negative CARs, one is not significantly different from zero, and the other is associated with a forced market closure that lasted 17 days. While this evidence is suggestive, it comports with previous research showing that political risk to markets declines with higher levels of democracy \citep{lehkonen2015democracy}.

@@

\autoref{fig:AR-Ven} shows CARs and 95\% confidence intervals for the 20 days surrounding the event. The abnormal return on the first trading day investors could react to the coup was +10\%. When Chavez was reinstated, the abnormal return was -8\%. The 0\% 10-day CAR preceding the coup implies that investors were completely unaware of the coup plot, increasing our confidence that the abnormal returns capture the true effect of the regime changes. This failed coup demonstrates a large influx of capital in response to the attempted overthrow of a socialist leader, and equally large capital flight following his reinstatement.  The large magnitudes and precision of these effects suggest that investors value transition to a pro-business government regime, regardless of how the regime change is achieved.

@@

Conventional theory suggests political instability causes capital flight. We show that this is not always the case. Unexpected changes in ruler virtually always increase volatility, but returns are not always negative. Domestic capital access can be given a boost when a new regime is expected to offer a more democratic or pro-business environment.

Despite their frequency, there is little evidence on the economic consequences of various types of political instability.  We show that resignations increase capital access on average, but assassinations and coups tend to cause capital flight. We also find evidence that capital markets tend to prefer democratic regime changes to authoritarian shifts, but that even democratically subversive coups can boost capital access if the instigators are clearly pro-business.

There are a number of avenues for future research. First, more research is needed to identify the pro and anti market characteristics of regime changes. Second, more work is needed to determine the extent to which financial market flows translate to broader economic development outcomes. It remains unclear whether the direction of the effects of different types of regime changes on outcomes such as economic growth, investment, debt, inflation, infant mortality, and years of schooling are consistent with their impact on financial flows, or if investor perceptions are at odds with certain development goals.
